---
title: "Point-by-point response analyses: mutation position and sequencing-quality checks"
author: "Francesco Gazzo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: cosmo
params:
  mpileup_summary_file: "mpileup_quality_summary_per_mutation.xlsx"
  phred_wide_file: "df_phred_wide.xlsx"
  theme_file: "ggplot.R"
---

# Aim of the analysis

This report documents the additional analyses performed for the point-by-point response.

The goal is to show, in a reproducible way, how the figures and statistical summaries were generated.

The report focuses on two main questions:

1. Are mutations spatially enriched around the integration site?
2. Do ALT-supporting bases show evidence of technical bias, such as proximity to read ends or reduced Phred quality?

# Setup

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 300
)

suppressPackageStartupMessages({
  library(openxlsx)
  library(dplyr)
  library(stringr)
  library(tidyr)
  library(purrr)
  library(broom)
  library(ggplot2)
  library(knitr)
})

# ----------------------------------------------------------
# Load custom ggplot theme, if available
# ----------------------------------------------------------

if (file.exists(params$theme_file)) {
  source(params$theme_file)
} else {
  message("ggplot.R not found. Using theme_bw() as fallback.")
  my_theme <- theme_bw(base_size = 12)
}

# ----------------------------------------------------------
# Output folders
# ----------------------------------------------------------

output_dir <- "point_by_point_outputs"
plot_dir <- file.path(output_dir, "plots")
table_dir <- file.path(output_dir, "tables")

dir.create(output_dir, showWarnings = FALSE)
dir.create(plot_dir, showWarnings = FALSE)
dir.create(table_dir, showWarnings = FALSE)

# ----------------------------------------------------------
# Dataset order and colors
# ----------------------------------------------------------

dataset_levels <- c(
  "Cdkn2a-/- LV.SF.LTR",
  "Cdkn2a-/- SIN.LV.PGK",
  "WT LV.SF.LTR",
  "WT SIN.LV.PGK"
)

dataset_colors <- c(
  "Cdkn2a-/- LV.SF.LTR" = "#7fb3d5",
  "Cdkn2a-/- SIN.LV.PGK" = "#FF8C00",
  "WT LV.SF.LTR" = "#8FBC8B",
  "WT SIN.LV.PGK" = "#D98880"
)
```

# Input files

This report uses two input files:

```{r input-files}
input_files <- c(
  params$mpileup_summary_file,
  params$phred_wide_file
)

input_files
```

```{r check-input-files}
missing_files <- input_files[!file.exists(input_files)]

if (length(missing_files) > 0) {
  stop(
    "Missing input file(s): ",
    paste(missing_files, collapse = ", "),
    ". Please place them in the working directory or update the YAML params."
  )
}
```

# Load data

```{r load-data}
df_raw <- read.xlsx(params$mpileup_summary_file)
df_phred <- read.xlsx(params$phred_wide_file)

data.frame(
  dataset = c("mpileup summary", "phred wide"),
  rows = c(nrow(df_raw), nrow(df_phred)),
  columns = c(ncol(df_raw), ncol(df_phred))
) %>%
  kable(caption = "Input dataset dimensions")
```

# Required columns

```{r required-columns}
required_raw_cols <- c(
  "ID_gene",
  "locus",
  "GenVec",
  "Mutation"
)

required_phred_cols <- c(
  "delta_ALT_minus_REF_dist_to_end",
  "delta_ALT_minus_REF_phred"
)

missing_raw_cols <- setdiff(required_raw_cols, names(df_raw))
missing_phred_cols <- setdiff(required_phred_cols, names(df_phred))

if (length(missing_raw_cols) > 0) {
  stop(
    "Missing column(s) in ",
    params$mpileup_summary_file,
    ": ",
    paste(missing_raw_cols, collapse = ", ")
  )
}

if (length(missing_phred_cols) > 0) {
  stop(
    "Missing column(s) in ",
    params$phred_wide_file,
    ": ",
    paste(missing_phred_cols, collapse = ", ")
  )
}
```

# 1. Spatial distribution of mutations relative to the integration site

## Rationale

This analysis evaluates where mutations are located relative to the integration site.

The integration-site coordinate is extracted from `ID_gene`.

The signed distance is calculated as:

```{r distance-definition, echo=FALSE}
data.frame(
  variable = "distance_from_is",
  definition = "locus - is_position"
) %>%
  kable(caption = "Definition of signed distance from integration site")
```

Interpretation:

- negative values indicate mutations upstream of the integration site;
- positive values indicate mutations downstream of the integration site;
- values close to zero indicate mutations close to the integration site.

## Build mutation-to-IS distance table

```{r build-distance-table}
df_distance <- df_raw %>%
  mutate(
    is_position = as.numeric(
      str_match(ID_gene, "chr[0-9A-Za-z]+_([0-9]+)_")[, 2]
    ),
    distance_from_is = locus - is_position,
    abs_distance_from_is = abs(distance_from_is),
    GenVec = factor(GenVec, levels = dataset_levels)
  ) %>%
  filter(
    !is.na(GenVec),
    !is.na(is_position),
    !is.na(locus),
    !is.na(distance_from_is)
  ) %>%
  separate(
    GenVec,
    into = c("Genotype", "Vector"),
    sep = " ",
    remove = FALSE,
    extra = "merge"
  ) %>%
  select(
    GenVec,
    Genotype,
    Vector,
    Mutation,
    ID_gene,
    is_position,
    locus,
    distance_from_is,
    abs_distance_from_is
  )

if (nrow(df_distance) == 0) {
  stop("No valid rows available after building df_distance.")
}

head(df_distance, 10) %>%
  kable(caption = "First rows of the mutation-to-IS distance table")
```

## Descriptive summary

```{r distance-summary}
distance_summary <- df_distance %>%
  group_by(GenVec, Genotype, Vector) %>%
  summarise(
    n_mutations = n(),
    mean_distance_from_is = mean(distance_from_is, na.rm = TRUE),
    median_distance_from_is = median(distance_from_is, na.rm = TRUE),
    sd_distance_from_is = sd(distance_from_is, na.rm = TRUE),
    median_abs_distance_from_is = median(abs_distance_from_is, na.rm = TRUE),
    iqr_abs_distance_from_is = IQR(abs_distance_from_is, na.rm = TRUE),
    min_distance_from_is = min(distance_from_is, na.rm = TRUE),
    max_distance_from_is = max(distance_from_is, na.rm = TRUE),
    pct_within_100_bp = mean(abs_distance_from_is <= 100, na.rm = TRUE) * 100,
    pct_within_1kb = mean(abs_distance_from_is <= 1000, na.rm = TRUE) * 100,
    .groups = "drop"
  )

write.csv(
  distance_summary,
  file.path(table_dir, "distance_from_is_summary.csv"),
  row.names = FALSE
)

distance_summary %>%
  kable(
    digits = 3,
    caption = "Summary of mutation distances from integration site"
  )
```

## Statistical test: is signed distance centered around zero?

For each dataset, a one-sample Wilcoxon signed-rank test is used to evaluate whether the signed distance distribution is centered around zero.

```{r safe-wilcox-function}
safe_wilcox_test <- function(x) {
  x <- x[!is.na(x)]

  if (length(x) < 3 || length(unique(x)) < 2) {
    return(tibble(
      statistic = NA_real_,
      p.value = NA_real_,
      method = "Wilcoxon signed-rank test",
      alternative = "two.sided",
      note = "Insufficient sample size or variability"
    ))
  }

  broom::tidy(
    wilcox.test(x, mu = 0, exact = FALSE)
  ) %>%
    mutate(note = NA_character_)
}
```

```{r distance-wilcox-tests}
distance_wilcox_results <- df_distance %>%
  group_by(GenVec) %>%
  summarise(values = list(distance_from_is), .groups = "drop") %>%
  mutate(test = map(values, safe_wilcox_test)) %>%
  select(-values) %>%
  unnest(test)

write.csv(
  distance_wilcox_results,
  file.path(table_dir, "distance_from_is_wilcox_tests.csv"),
  row.names = FALSE
)

distance_wilcox_results %>%
  kable(
    digits = 4,
    caption = "Wilcoxon signed-rank tests on signed distance from IS"
  )
```

## Statistical test: compare absolute distance across datasets

The Kruskal-Wallis test is used to compare the absolute mutation distance from the integration site across datasets.

```{r distance-kruskal}
distance_kruskal_result <- broom::tidy(
  kruskal.test(abs_distance_from_is ~ GenVec, data = df_distance)
)

write.csv(
  distance_kruskal_result,
  file.path(table_dir, "absolute_distance_from_is_kruskal_test.csv"),
  row.names = FALSE
)

distance_kruskal_result %>%
  kable(
    digits = 4,
    caption = "Kruskal-Wallis test on absolute distance from IS"
  )
```

## Plot: signed distance from integration site

```{r plot-distance-histogram}
p_distance_hist <- ggplot(
  df_distance,
  aes(x = distance_from_is, fill = GenVec)
) +
  geom_histogram(
    bins = 50,
    alpha = 0.7,
    position = "identity"
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    linewidth = 0.4
  ) +
  facet_grid(Genotype ~ Vector) +
  scale_fill_manual(values = dataset_colors, drop = FALSE) +
  my_theme +
  labs(
    title = "Spatial distribution of mutations relative to the integration site",
    x = "Signed distance from IS, bp",
    y = "Number of mutations"
  ) +
  theme(
    legend.position = "none",
    strip.background = element_blank()
  )

p_distance_hist

ggsave(
  filename = file.path(plot_dir, "distance_from_is_histogram.png"),
  plot = p_distance_hist,
  width = 8,
  height = 5,
  dpi = 300
)

ggsave(
  filename = file.path(plot_dir, "distance_from_is_histogram.pdf"),
  plot = p_distance_hist,
  width = 8,
  height = 5
)
```

## Plot: absolute distance from integration site

```{r plot-absolute-distance-boxplot}
p_abs_distance_boxplot <- ggplot(
  df_distance,
  aes(x = GenVec, y = abs_distance_from_is, fill = GenVec)
) +
  geom_boxplot(outlier.alpha = 0.35) +
  scale_y_log10() +
  scale_fill_manual(values = dataset_colors, drop = FALSE) +
  my_theme +
  labs(
    title = "Absolute distance between mutations and the integration site",
    x = NULL,
    y = "Absolute distance from IS, bp, log10 scale"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 35, hjust = 1)
  )

p_abs_distance_boxplot

ggsave(
  filename = file.path(plot_dir, "absolute_distance_from_is_boxplot.png"),
  plot = p_abs_distance_boxplot,
  width = 8,
  height = 5,
  dpi = 300
)

ggsave(
  filename = file.path(plot_dir, "absolute_distance_from_is_boxplot.pdf"),
  plot = p_abs_distance_boxplot,
  width = 8,
  height = 5
)
```

# 2. Read-end and Phred-quality checks

## Rationale

This analysis evaluates whether ALT-supporting bases show evidence of technical bias compared with REF-supporting bases.

The two main variables are:

- `delta_ALT_minus_REF_dist_to_end`
- `delta_ALT_minus_REF_phred`

Interpretation:

- `delta_ALT_minus_REF_dist_to_end < 0`: ALT-supporting bases are closer to the nearest read end than REF-supporting bases.
- `delta_ALT_minus_REF_dist_to_end > 0`: ALT-supporting bases are farther from the nearest read end than REF-supporting bases.
- `delta_ALT_minus_REF_phred < 0`: ALT-supporting bases have lower mean Phred quality than REF-supporting bases.
- `delta_ALT_minus_REF_phred > 0`: ALT-supporting bases have higher mean Phred quality than REF-supporting bases.

## Build quality table

```{r build-quality-table}
df_quality <- df_phred %>%
  mutate(
    delta_ALT_minus_REF_dist_to_end =
      as.numeric(delta_ALT_minus_REF_dist_to_end),
    delta_ALT_minus_REF_phred =
      as.numeric(delta_ALT_minus_REF_phred)
  ) %>%
  filter(
    !is.na(delta_ALT_minus_REF_dist_to_end),
    !is.na(delta_ALT_minus_REF_phred)
  )

if ("GenVec" %in% names(df_quality)) {
  df_quality <- df_quality %>%
    mutate(GenVec = factor(GenVec, levels = dataset_levels))
}

if (nrow(df_quality) == 0) {
  stop("No valid rows available after building df_quality.")
}

head(df_quality, 10) %>%
  kable(caption = "First rows of the quality table")
```

## Descriptive summary

```{r quality-summary}
quality_summary <- df_quality %>%
  summarise(
    n_snp_mutations = n(),

    mean_delta_read_end_distance =
      mean(delta_ALT_minus_REF_dist_to_end, na.rm = TRUE),

    median_delta_read_end_distance =
      median(delta_ALT_minus_REF_dist_to_end, na.rm = TRUE),

    iqr_delta_read_end_distance =
      IQR(delta_ALT_minus_REF_dist_to_end, na.rm = TRUE),

    pct_alt_closer_to_read_end =
      mean(delta_ALT_minus_REF_dist_to_end < 0, na.rm = TRUE) * 100,

    mean_delta_phred =
      mean(delta_ALT_minus_REF_phred, na.rm = TRUE),

    median_delta_phred =
      median(delta_ALT_minus_REF_phred, na.rm = TRUE),

    iqr_delta_phred =
      IQR(delta_ALT_minus_REF_phred, na.rm = TRUE),

    pct_alt_lower_phred =
      mean(delta_ALT_minus_REF_phred < 0, na.rm = TRUE) * 100
  )

write.csv(
  quality_summary,
  file.path(table_dir, "quality_delta_summary.csv"),
  row.names = FALSE
)

quality_summary %>%
  kable(
    digits = 4,
    caption = "Global ALT-minus-REF read-end and Phred-quality summaries"
  )
```

## Optional summary by dataset

This section is generated only if `GenVec` is present in `df_phred_wide.xlsx`.

```{r quality-summary-by-dataset}
if ("GenVec" %in% names(df_quality) && any(!is.na(df_quality$GenVec))) {

  quality_summary_by_dataset <- df_quality %>%
    filter(!is.na(GenVec)) %>%
    group_by(GenVec) %>%
    summarise(
      n_snp_mutations = n(),

      median_delta_read_end_distance =
        median(delta_ALT_minus_REF_dist_to_end, na.rm = TRUE),

      pct_alt_closer_to_read_end =
        mean(delta_ALT_minus_REF_dist_to_end < 0, na.rm = TRUE) * 100,

      median_delta_phred =
        median(delta_ALT_minus_REF_phred, na.rm = TRUE),

      pct_alt_lower_phred =
        mean(delta_ALT_minus_REF_phred < 0, na.rm = TRUE) * 100,

      .groups = "drop"
    )

  write.csv(
    quality_summary_by_dataset,
    file.path(table_dir, "quality_delta_summary_by_dataset.csv"),
    row.names = FALSE
  )

  quality_summary_by_dataset %>%
    kable(
      digits = 4,
      caption = "ALT-minus-REF read-end and Phred-quality summaries by dataset"
    )
}
```

## Statistical tests on quality deltas

A one-sample Wilcoxon signed-rank test is used to evaluate whether each ALT-minus-REF delta distribution is centered around zero.

```{r quality-wilcox-tests}
read_end_wilcox <- safe_wilcox_test(
  df_quality$delta_ALT_minus_REF_dist_to_end
) %>%
  mutate(variable = "delta_ALT_minus_REF_dist_to_end", .before = 1)

phred_wilcox <- safe_wilcox_test(
  df_quality$delta_ALT_minus_REF_phred
) %>%
  mutate(variable = "delta_ALT_minus_REF_phred", .before = 1)

quality_wilcox_results <- bind_rows(
  read_end_wilcox,
  phred_wilcox
)

write.csv(
  quality_wilcox_results,
  file.path(table_dir, "quality_delta_wilcox_tests.csv"),
  row.names = FALSE
)

quality_wilcox_results %>%
  kable(
    digits = 4,
    caption = "Wilcoxon signed-rank tests on ALT-minus-REF quality deltas"
  )
```

## Plot: ALT-minus-REF distance from nearest read end

```{r plot-read-end-distance}
p_read_end <- ggplot(
  df_quality,
  aes(x = delta_ALT_minus_REF_dist_to_end)
) +
  geom_histogram(bins = 60) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_bw() +
  labs(
    x = "ALT - REF relative distance from nearest read end",
    y = "Number of SNP mutations",
    title = "No enrichment of ALT-supporting bases near read ends"
  )

p_read_end

ggsave(
  filename = file.path(plot_dir, "alt_minus_ref_read_end_distance.png"),
  plot = p_read_end,
  width = 8,
  height = 5,
  dpi = 300
)

ggsave(
  filename = file.path(plot_dir, "alt_minus_ref_read_end_distance.pdf"),
  plot = p_read_end,
  width = 8,
  height = 5
)
```

## Plot: ALT-minus-REF Phred quality

```{r plot-phred-quality}
p_phred <- ggplot(
  df_quality,
  aes(x = delta_ALT_minus_REF_phred)
) +
  geom_histogram(bins = 60) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_bw() +
  labs(
    x = "ALT - REF mean Phred quality",
    y = "Number of SNP mutations",
    title = "ALT-supporting bases show a modest Phred reduction"
  )

p_phred

ggsave(
  filename = file.path(plot_dir, "alt_minus_ref_phred_quality.png"),
  plot = p_phred,
  width = 8,
  height = 5,
  dpi = 300
)

ggsave(
  filename = file.path(plot_dir, "alt_minus_ref_phred_quality.pdf"),
  plot = p_phred,
  width = 8,
  height = 5
)
```

# 3. Optional stratified quality plots

These plots are generated only if the `GenVec` column is available in `df_phred_wide.xlsx`.

```{r optional-stratified-quality-plots}
if ("GenVec" %in% names(df_quality) && any(!is.na(df_quality$GenVec))) {

  p_read_end_by_dataset <- ggplot(
    df_quality,
    aes(x = delta_ALT_minus_REF_dist_to_end, fill = GenVec)
  ) +
    geom_histogram(
      bins = 60,
      alpha = 0.7,
      position = "identity"
    ) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    facet_wrap(~ GenVec, scales = "free_y") +
    scale_fill_manual(values = dataset_colors, drop = FALSE) +
    theme_bw() +
    labs(
      x = "ALT - REF relative distance from nearest read end",
      y = "Number of SNP mutations",
      title = "Read-end distance delta by dataset"
    ) +
    theme(legend.position = "none")

  p_read_end_by_dataset

  ggsave(
    filename = file.path(plot_dir, "alt_minus_ref_read_end_distance_by_dataset.png"),
    plot = p_read_end_by_dataset,
    width = 9,
    height = 6,
    dpi = 300
  )

  p_phred_by_dataset <- ggplot(
    df_quality,
    aes(x = delta_ALT_minus_REF_phred, fill = GenVec)
  ) +
    geom_histogram(
      bins = 60,
      alpha = 0.7,
      position = "identity"
    ) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    facet_wrap(~ GenVec, scales = "free_y") +
    scale_fill_manual(values = dataset_colors, drop = FALSE) +
    theme_bw() +
    labs(
      x = "ALT - REF mean Phred quality",
      y = "Number of SNP mutations",
      title = "Phred-quality delta by dataset"
    ) +
    theme(legend.position = "none")

  p_phred_by_dataset

  ggsave(
    filename = file.path(plot_dir, "alt_minus_ref_phred_quality_by_dataset.png"),
    plot = p_phred_by_dataset,
    width = 9,
    height = 6,
    dpi = 300
  )
}
```

# 4. Output files

The report also saves the main output tables and plots to disk.

```{r output-files}
list.files(output_dir, recursive = TRUE) %>%
  data.frame(file = .) %>%
  kable(caption = "Files generated by this report")
```

# 5. Reproducibility

```{r session-info}
sessionInfo()
```